Rheology of Cubic Blue Phases

### 2012-08-14 trunk/papers/bprheo/bprheo.tex
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1749 40876441-3496-0410-a613-a2fca35b2ae2

\newcommand{\e}[1]{\times10^{#1}}

@@

\begin{abstract}
bla
\end{abstract}

@@

Simple shear flow is imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction.
We applied a correction for spurious momentum currents that emerge due to interpolations accross 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$2.44\cdot 10^{-6}$ to $6.3\cdot10^{-4}$ in LB units.
The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
along the y- and z-axis.
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
is equal to about $10^8$~Pa.
More details about the conversion from LBU to SI units can be found in
\cite{Henrich:2011a,Henrich:2010b}.

@@

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
of shear-induced transformations in the steady state at 
$\dot{\gamma}=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
The boundary conditions are applied flow takes place in such as way that network moves 
to the right in the upper half and to the left in the lower half.}
\label{bp2-med}
\end{figure*}

Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
the gradient direction in the direction of vorticity.
Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
geometries~\cite{Henrich:2012b} the movement is steady. It happens to be in such a way that the 
positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
making the network travel along the vorticity direction.

Fig. \ref{bp2-velo} shows the disclination network at a characteristic time step between reconnecting and another breakup.
Superimposed are vectors which represent the velocity projected onto a plain perpendicular to the flow direction.
This projection makes it possible to get rid of the dominating velocity component $v_x$ and permits visualisation the 
two much smaller components $v_y$ and $v_z$.
The magnitude of the minor components is typically in the range of a few percent of the primary flow component.
Characteristic bands oriented along the vorticity direction form, which are normally hidden behind the primary flow in x-direction.
When the helicity of the underlying cholesteric phase is changed from a left- to a righthanded helix the sign of $v_z$ 
and the sense of motion also changes.
Further quantitative evidence for a direct link between flow direction and helicity can be found by time-averaging over
individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
All values for the two runs with inverted helicity are identical apart from a change of sign of the z-components.

@@

It is interesting to study the flow behaviour from a more quantitative point of view by looking at the shear stress dependence on flow rate and time.

The effective viscosity $\eta_{eff}$ over time is shown in Fig. \ref{fig4}. 
For all apart from the highest flow rate it oscillates sinusoidally, exhibiting symmetric minima and maxima and decreasing mean values, indicating slightly shear-thinning behaviour.
The period reflects the time it takes for the disclination network to break up and reconnect as shown in Fig. \ref{fig11}.
However, due to the motion of the network in z-direction it turns out that it takes four periods altogether for the network to move one unit cell along the direction of vorticity and to complete one cycle.
The inset shows the deviatoric part of the chemical stress tensor as a function of shear rate.
The dependence is that of a power law $\sigma_{eff, xy}=a \dot{\gamma}^b$ with $a=0.34(6116), b=0.95(2654)$ for low and intermediate shear rates. 
The error bars indicate the maximum and minimum stress during one cycle.
The idependence is almost linear and gives further evidence of the remarkably small degree of shear-thinning and also reflected in the very regular and undisturbed affine transformation of the disclination network.

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ of BPII over time (left) and strain (right). The inset shows the flow curve $\sigma_{chem}(\dot{\gamma})$.}
\label{bp2-rheo}
\end{figure}

Above a critical shear rate, which is $\dot{\gamma}=3.1-6.2\cdot 10^{-4}$ in our case, the rotational diffusion coefficient $\Gamma$ in Eq. \ref{eqn2} is too small and the disclination network adopts a completely different conformation in the steady state.
Fig. \ref{fig5} shows the director field and the position of disclination plane.
The degree of twist is the same as that of the underlying cholesteric, but the orientation of the director on either side of the planes differs by almost 90 degrees, giving rise to four distinct layers, which are separated by disclination planes.
Even in this flow-induced state the disclinations move and retain the above described sense of motion in the direction of vorticity.

\begin{figure}[h]
\caption{BPII at low shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom). The configuration is a travelling helical wave along the vorticity direction. The configuration is translationally invariant in flow and in gradient direction. The colour code indicates the orientation in x-y direction.}
\label{bp2-low}
\end{figure}

\begin{figure}[h]
\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom). The configuration is a travelling helical wave along the vorticity direction. The configuration is translationally invariant in flow and in gradient direction. The colour code indicates the orientation in x-y direction.}
\label{bp2-high}
\end{figure}

@@

The flow behaviour of BPI deviates significantly from that of BPII in a number of points.
First of all it takes considerably longer to reach the steady state with periodically recurring patterns.
This might be because the disclination network of BPI undergoes much more complex flow-induced transformations than that of BPII.
Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig11}, but this time seen along the direction of flow.
A similar affine transformation in flow-gradient plane takes place, which is however obscurred by the transformations shown here. 
The initial and final state in this sequence correspond to half a cycle. 
This can be easily verified by comparing the position of the disclination network at the left or right boundary.
BPI and BPII have exactly the same recurrence period, although it is seemingly four times shorter if the travelling of the network in z-direction is not taken into account.

@@

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
\label{bp1-spec}
\end{figure}

@@

\begin{table*}
\begin{tabular}{|c|| c | c| c || c |c |c||c| c| c||c| c| c|}
\hline
$\dot{\gamma}$ & $\tau$ & $\kappa$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI \\
\hline
1.95 & -0.5 & 1.0 &0.1388 & -2.49 &2.50 &4.51 &-124.65 &124.82 &3.62 &-1.62 &1.89 &3.51 \\
3.91 & -0.5 & 1.0 &0.1388 & -3.31 &3.35 &4.21 &-248.69 &248.74 &3.90 &-2.88 &2.56 &4.39 \\
7.81 & -0.5 & 1.0 &0.1388 &-6.49 &6.56 &7.99 &-497.22 &496.16 &4.46 &-7.46 &5.31 &6.81 \\ 
15.63 & -0.5 & 1.0 &0.1388 &-3.08 &3.17 &10.49 &-992.46 &993.18 &10.36 &-3.57 &2.87 &10.54 \\
15.63 & -0.5 & 1.0 &-0.1388 &-3.08 &3.17 &10.49 &-992.46 &993.18 &10.36 &-2.87 &3.57 &10.54 \\
31.25 & -0.5 & 1.0 &0.1388 &-4.08 &4.46 &15.94 & -1988.36 &1986.71 &16.85 &-12.16 &11.37 &19.38\\
62.50 & -0.5 & 1.0 &0.1388 & -62.63 & 62.13 & 24.68 & -4039.41 &3995.3  & 20.57 &-110.76  &73.52 & 33.26 \\
\hline
BPII \\
\hline
1.95 & -0.5 & 2.0 & 0.1963 & -1.09 &1.07 & 0.76 &  -62.04  &61.60 & 0.97 & -1.55 &1.64 & 0.81\\
3.91 & -0.5 & 2.0 & 0.1963&-1.66 &1.70 & 1.45 &-123.93 &123.18 & 1.76 &-2.73 &3.09 &1.47\\
7.81 &  -0.5 & 2.0 &0.1963& -2.62 &2.71 & 2.67 &-247.67 &246.45 & 3.20 &-4.77 &5.78 &2.74\\
15.63 & -0.5 & 2.0 & 0.1963&-3.13 &3.31 &4.24 &-494.85 &493.09 & 5.64 &-7.66 &10.00 &4.33\\
15.63 & -0.5 & 2.0 &-0.1963& -3.13 &3.31 &4.24 &-494.85 &493.09 & 5.64 &-10.00 &7.66 &4.33\\
31.25 & -0.5 & 2.0 &0.1963 &-1.79 &1.82 &5.86 &-988.73 &985.93 &8.96  &-11.04 &14.39 &6.35\\
62.50 & -0.5 & 2.0 &0.1963 & -1.20 & 1.20 & 0.38 & -1969.41  & 1969.41 & 3.07 &-1.48 &0.96 &0.38 \\
\hline
\end{tabular}
\caption{Time-averaged velocity components in shear flow: Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
\label{tab1}}
\end{table*}
%\section{}

>>>>

\newcommand{\e}[1]{\times10^{#1}}
\newcommand{\gd}{\dot{\gamma}}

@@

\begin{abstract}
We study the behaviour of cubic blue phases under shear flow
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under shear, leading to an oscillatory stress response in time.
\end{abstract}

@@

Simple shear flow is imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction.
We applied a correction for spurious momentum currents that emerge due to interpolations accross 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.3\times\e{-4}$ in LB units.
The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
along the y- and z-axis.
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
is equal to about $10^8$~Pa.
More details about the conversion from LBU to SI units can be found in
\cite{Henrich:2011a,Henrich:2010b}.

@@

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
of shear-induced transformations in the steady state at 
$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
The boundary conditions are applied flow takes place in such as way that network moves 
to the right in the upper half and to the left in the lower half.}
\label{bp2-med}
\end{figure*}

Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
the gradient direction in the direction of vorticity.
Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
making the network travel along the z-direction.
The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.

Fig. \ref{bp2-velo} shows the disclination network at a characteristic time step between reconnecting and another breakup.
Superimposed are vectors which represent the velocity projected onto a plain perpendicular to the flow direction.
This projection makes it possible to get rid of the dominating velocity component $v_x$ and permits visualisation the 
two much smaller components $v_y$ and $v_z$.
The magnitude of the minor components is typically in the range of a few percent of the primary flow component.
Characteristic bands oriented along the vorticity direction form, which are normally hidden behind the primary flow in x-direction.
When the helicity of the underlying cholesteric phase is changed from a left- to a righthanded helix the sign of $v_z$ 
and the sense of motion also changes.
Further quantitative evidence for a direct link between flow direction and helicity can be found by time-averaging over
individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
All values for the two runs with inverted helicity are identical apart from a change of sign in the z-components.
There is only a slight imbalance between the maximum and minimum velocity in z-direction.
Hence the movement of the network in vorticity direction is predominantly permeative.
Note that the imbalance is exaclty retained if the helicity is inverted.

@@

If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation under shear that we
descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
shear rates \cite{Rey:1996a, Rey:1996b}.

In our scenario the break-up occurs for shear rates between $\gd=3.12\e{-4}$ and $6.25\e{-4}$.
However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
Due to a lack of computational resources we have not investigated these aspects any further.

\begin{figure}[h]
\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom). The images show 1D-cuts as the configuration is translationally invariant in flow and gradient direction. The configuration is a travelling helical wave along the vorticity or z-direction. The colour code indicates the orientation in x-y direction.}
\label{bp2-high}
\end{figure}

A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
the picture on the right shows the same data versus accummulated total strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is due to the periodic break-up and reconnection of the network under shear flow.

The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
The error bars indicate the maximum and minimum stresses that occur during one cycle.
The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
a very good approximation, exibiting a remarkably small degree of shear-thinning.
This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp2-rheo}
\end{figure*}

@@

The flow behaviour of BPI at low shear rates is is similar deviates significantly from that of BPII in a number of points.

First of all it takes considerably longer to reach the steady state with periodically recurring patterns.
This might be because the disclination network of BPI undergoes much more complex flow-induced transformations than that of BPII.
Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig11}, but this time seen along the direction of flow.
A similar affine transformation in flow-gradient plane takes place, which is however obscurred by the transformations shown here. 
The initial and final state in this sequence correspond to half a cycle. 
This can be easily verified by comparing the position of the disclination network at the left or right boundary.
BPI and BPII have exactly the same recurrence period, although it is seemingly four times shorter if the travelling of the network in z-direction is not taken into account.

@@

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
\label{bp1-spec}
\end{figure}

@@

\begin{table*}
\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
\hline
$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI \\
\hline
0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
\hline
BPII \\
\hline
0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
\hline
\end{tabular}
\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
\label{tab1}}
\end{table*}
%\section{}
